---
title: "NVSS natality"
author: "Michele Santacatterina"
date: "2025-11-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(vroom)
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
```

Assumptions

- Use 2016+ public-use NVSS (smoking by trimester available nationally). Variables from the 2022/2023 User Guides. 

- Restrict to live, singleton, term births if desired; exclude implausible weights per the Guide.

- Exposure: any smoking during pregnancy, or trimester-specific counts (convert to intensity bins).

- Outcomes: continuous birth weight (DBWT) and LBW (<2500 g).

- Confounders: maternal age (MAGER), education (MEDUC 2003-rev scale), race/ethnicity (MRACE6 + Hispanic), marital status (DMAR), parity/live-birth order (LBO_REC or TBO), prenatal care timing (PRECARE), payment source (PAY), plurality, state and year fixed effects.

```{r}

# paths
PATH_DIR <- "~/Documents/Projects/kernel-blb/data"  # change if needed
DCT_URL  <- "https://data.nber.org/nvss/natality/programs/dct/natality2023.dct"

# locate the TXT
txt_candidates <- list.files(PATH_DIR, pattern="\\.txt$", full.names=TRUE, ignore.case=TRUE)
stopifnot(length(txt_candidates) >= 1)
TXT <- txt_candidates[1]

# load and parse the 2023 .dct
DCT_FILE <- file.path(tempdir(), "natality2023.dct")
download.file(DCT_URL, DCT_FILE, mode="wb", quiet=TRUE)
dct <- read_lines(DCT_FILE)

rx <- "_column\\((\\d+)\\)\\s+([a-z0-9]+)\\s+([A-Za-z0-9_]+)\\s+%([0-9]+)"
m  <- str_match(dct, rx)
dct_df <- data.frame(
  start = as.integer(m[,2]),
  type  = m[,3],
  name  = m[,4],
  width = as.integer(m[,5]),
  stringsAsFactors = FALSE
)
dct_df <- dct_df[!is.na(dct_df$start), ]
dct_df <- dct_df[order(dct_df$start), ]
dct_df$end <- dct_df$start + dct_df$width - 1
stopifnot(!is.na(TXT), file.exists(TXT))


```


```{r, include=T}
# Byte positions (inclusive) from your 2023 .dct snippet
# dob_yy 9–12; mager 75–76; mrace6 107; mhisp_r 115; dmar 120; meduc 124;
# lbo_rec 179; precare 224–225; cig_1 255–256; cig_2 257–258; cig_3 259–260;
# dplural 454; dbwt 504–507; pay 435
pos <- fwf_positions(
  start = c(   9,  75, 107, 115, 120, 124, 179, 224, 255, 257, 259, 454, 504, 435),
  end   = c(  12,  76, 107, 115, 120, 124, 179, 225, 256, 258, 260, 454, 507, 435),
  col_names = c("dob_yy","mager","mrace6","mhisp_r","dmar","meduc","lbo_rec",
                "precare","cig_1","cig_2","cig_3","dplural","dbwt","pay")
)

# Read only those columns (fast, robust to .dct quirks)
raw <- vroom_fwf(TXT, col_positions = pos, progress = TRUE, altrep = TRUE)

# Minimal hygiene and variable construction
nv <- raw |>
  mutate(across(everything(), \(x) na_if(trimws(as.character(x)), ""))) |>
  transmute(
    # outcomes
    dbwt    = as.integer(ifelse(dbwt == "9999", NA, dbwt)),
    lbw     = as.integer(dbwt < 2500),

    # exposure
    cig1    = as.integer(ifelse(cig_1 %in% c("", "99"), NA, cig_1)),
    cig2    = as.integer(ifelse(cig_2 %in% c("", "99"), NA, cig_2)),
    cig3    = as.integer(ifelse(cig_3 %in% c("", "99"), NA, cig_3)),
    any_smoke = as.integer((coalesce(cig1, 0) > 0) | (coalesce(cig2, 0) > 0) | (coalesce(cig3, 0) > 0)),

    # confounders
    age     = as.integer(ifelse(mager %in% c("", "99"), NA, mager)),
    educ    = factor(ifelse(meduc %in% c("", "9"), NA, meduc)),
    race6   = factor(ifelse(mrace6 %in% c("", "9"), NA, mrace6)),
    hisp    = factor(ifelse(mhisp_r %in% c("", "9"), NA, mhisp_r)),
    married = factor(ifelse(dmar %in% c("", "9"), NA, dmar)),
    parity  = as.integer(ifelse(lbo_rec %in% c("", "9"), NA, lbo_rec)),
    precare = as.integer(ifelse(precare %in% c("", "99"), NA, precare)),
    pay     = factor(ifelse(pay %in% c("", "9"), NA, pay)),
    plural1 = as.integer(dplural) == 1,
    year    = as.integer(dob_yy)
  ) |>
  filter(plural1, dbwt >= 350, dbwt <= 6000) |>
  drop_na(dbwt, age, educ, race6, hisp, married, parity, precare, pay, any_smoke, year)

```


```{r}

#Augmented Inverse Probability Weighting (EIF based standard error)

data <- nv
n <- nrow(data)

e  <- predict(glm(any_smoke ~ age + I(educ) + I(race6) + I(hisp) +
    married + parity + precare + I(pay) , data = data, family = "binomial"),
              type = "response")
m0 <- predict(glm(dbwt ~ age + I(educ) + I(race6) + I(hisp) +
    married + parity + precare + I(pay), subset = (any_smoke == 0), data = data), newdata = data)

m1 <- predict(glm(dbwt ~ age + I(educ) + I(race6) + I(hisp) +
    married + parity + precare + I(pay), subset = (any_smoke == 1), data = data), newdata = data)

# trunc_bounds <- quantile(e, probs = c(0.1, 0.9), na.rm = TRUE)
# e <- pmin(pmax(e, trunc_bounds[1]), trunc_bounds[2])

phi1 <- (data$any_smoke * data$dbwt - (data$any_smoke - e) * m1)/e
phi0 <- ( (1 - data$any_smoke) * data$dbwt - (data$any_smoke - e) * m0 )/(1-e)

mu_hat <- mean(phi1) - mean(phi0)
se_hat <- sqrt( var(phi1)/n + var(phi0)/n )

ate_PE_aipw_eif <- mu_hat
ate_SE_aipw_eif <- se_hat
ate_CI_aipw_eif <- c(ate_PE_aipw_eif-qnorm(0.975)*ate_SE_aipw_eif, ate_PE_aipw_eif+qnorm(0.975)*ate_SE_aipw_eif)

W <- ate_PE_aipw_eif/ate_SE_aipw_eif
ate_PV_aipw_eif <- 2*pnorm(-abs(W))

ate_aipw_eif <- data.frame("AIPW (EIF)",ate_PE_aipw_eif,ate_SE_aipw_eif,t(ate_CI_aipw_eif),ate_PV_aipw_eif)
colnames(ate_aipw_eif) <-  c("Method","ATE estimate","SE","2.5 %","97.5 %","Pvalue")
ate_aipw_eif

naive <- mean(data$dbwt[data$any_smoke==1]) - mean(data$dbwt[data$any_smoke==0])
naive

print(summary(phi0))
```

# ```{r}
# library(glmnet)
# 
# #Augmented Inverse Probability Weighting (EIF based standard error)
# 
# data <- nv
# n <- nrow(data)
# 
# ## Cross-fitted penalized logistic regression for propensity scores
# X <- model.matrix(any_smoke ~ age + I(educ) + I(race6) + I(hisp) +
#                     married + parity + precare + I(pay),
#                   data = data)[, -1]
# 
# y <- data$any_smoke
# if (is.factor(y)) {
#   y <- as.numeric(y) - 1
# }
# 
# K <- 5
# set.seed(123)
# folds <- sample(rep(1:K, length.out = n))
# e <- numeric(n)
# 
# for (k in 1:K) {
#   train_idx <- which(folds != k)
#   test_idx  <- which(folds == k)
#   fit <- cv.glmnet(
#     x = X[train_idx, , drop = FALSE],
#     y = y[train_idx],
#     family = "binomial",
#     alpha  = 1
#   )
#   e[test_idx] <- as.numeric(
#     predict(fit, newx = X[test_idx, , drop = FALSE],
#             s = "lambda.min", type = "response")
#   )
# }
# 
# m0 <- predict(glm(dbwt ~ age + I(educ) + I(race6) + I(hisp) +
#                     married + parity + precare + I(pay),
#                   subset = (any_smoke == 0), data = data),
#               newdata = data)
# 
# m1 <- predict(glm(dbwt ~ age + I(educ) + I(race6) + I(hisp) +
#                     married + parity + precare + I(pay),
#                   subset = (any_smoke == 1), data = data),
#               newdata = data)
# 
# # trunc_bounds <- quantile(e, probs = c(0.1, 0.9), na.rm = TRUE)
# # e <- pmin(pmax(e, trunc_bounds[1]), trunc_bounds[2])
# 
# phi1 <- (data$any_smoke * data$dbwt - (data$any_smoke - e) * m1)/e
# phi0 <- ((1 - data$any_smoke) * data$dbwt - (data$any_smoke - e) * m0)/(1 - e)
# 
# mu_hat <- mean(phi1) - mean(phi0)
# se_hat <- sqrt(var(phi1)/n + var(phi0)/n)
# 
# ate_PE_aipw_eif <- mu_hat
# ate_SE_aipw_eif <- se_hat
# ate_CI_aipw_eif <- c(
#   ate_PE_aipw_eif - qnorm(0.975) * ate_SE_aipw_eif,
#   ate_PE_aipw_eif + qnorm(0.975) * ate_SE_aipw_eif
# )
# 
# W <- ate_PE_aipw_eif/ate_SE_aipw_eif
# ate_PV_aipw_eif <- 2 * pnorm(-abs(W))
# 
# ate_aipw_eif <- data.frame(
#   "AIPW (EIF)", ate_PE_aipw_eif, ate_SE_aipw_eif,
#   t(ate_CI_aipw_eif), ate_PV_aipw_eif
# )
# colnames(ate_aipw_eif) <- c("Method", "ATE estimate", "SE", "2.5 %", "97.5 %", "Pvalue")
# ate_aipw_eif
# print(phi0)
# 
# ```
